--- 
title: amen
author: Peter Hoff
date: 2020/12-01 
output:
  md_document:
    variant: markdown_github
---


## amen

#### Additive and multiplicative effects network models



[Package website](https://pdhoff.github.io/amen/)


```{r,echo=FALSE}
library(amen)
```

```{r fitConflict,echo=FALSE,cache=TRUE,fig.keep='none',results='hide'}
data(IR90s)
Y<-1*(IR90s$dyadvars[,,1]>0)
fitC<-ame(Y,family="bin",R=2,nscan=5000) 
```


```{r,echo=FALSE}
circplot(IR90s$dyadvars[,,1],U=fitC$U,V=fitC$V,pscale=1.15,mscale=.3,jitter=.3)
```


#### About
Additive and multiplicative effects network models (AMEN models)
provide a statistical modeling framework for dyadic network and 
relational data, built upon familiar data analysis tools such as 
linear regression, random effects models and matrix decompositions. 
The `amen` package provides Bayesian model fitting algorithms
for AMEN models, and accommodates a variety of types of 
network relations, 
including continuous, binary and ordinal dyadic variables. 

The basic AMEN model is of the form 
$y_{i,j} \sim  \beta^\top x_{i,j} + u_i^\top v_j + a_i +b_j + \epsilon_{i,j}$
where 

* $y_{i,j}$ is the observed dyadic variable being modeled
and $x_{i,j}$ is an observed vector of regressors;

* $a_i + b_j +  \epsilon_{i,j}$ is an  additive random effects term  that describes sender and receiver variance (such as outdegree and indegree 
heterogeneity) and dyadic correlation; 

* $u_i^\top v_j$ is a multiplicative random effects term 
that describes third-order 
dependence patterns (such as transitivity and clustering) and can be estimated 
and analyzed to uncover low-dimensional  structure in the network. 

#### Installation


```{r,eval=FALSE}
# Current version on GitHub
devtools::install_github("pdhoff/amen") 

# CRAN-approved version on CRAN
install.packages("amen")
```


#### Documentation and citation

A tutorial article and many data analysis examples are available 
via the [vignette](https://github.com/pdhoff/amen/blob/master/inst/doc/amen.pdf). Please cite this as 

Hoff, P.D. (2015) "Dyadic data analysis with *amen*". arXiv:1506.08237. 

A review article that provides some mathematical details and derivations 
is available on [arXiv](https://arxiv.org/abs/1807.08038). Please cite this

Hoff, P.D. (2018) "Additive and multiplicative effects network models". arXiv:1807.08038. 

The first version of the AMEN model appeared in 

Hoff, P.D. (2005) "Bilinear mixed-effects models for dyadic data".
JASA 100(469) 286-295. 

That version restricted the multiplicative sender and receiver effects to be equal ($u_i=v_i$). The AMEN model in its current form does not have this 
restriction. The current AMEN model first appeared in 

Hoff, P.D., Fosdick, B.K., Volfovsky, A. and Stovel, K. (2013) "Likelihoods for fixed rank nomination networks". Network Science, 1(3):253–277. 


#### Some examples

* [Modeling a binary outcome](articles/binary_demo.html)

* [DIY Modeling of a binary outcome](articles/diy_binary_demo.html)

* [DIY Poisson regression](articles/diy_Poisson_demo.html)


