## amen

#### Additive and multiplicative effects network models



[Package website](https://pdhoff.github.io/amen/)


```{r,echo=FALSE}
library(amen)
```

```{r fitConflict,echo=FALSE,cache=TRUE,fig.keep='none',results='hide'}
data(IR90s)
Y<-1*(IR90s$dyadvars[,,1]>0)
fitC<-ame(Y,family="bin",R=2,nscan=5000) 
```


```{r,echo=FALSE}
circplot(IR90s$dyadvars[,,1],U=fitC$U,V=fitC$V,pscale=1.15,mscale=.3,jitter=.3)
```


#### About
Additive and multiplicative effects (AME) models 
provide a statistical modeling framework for dyadic network and 
relational data, built upon familiar data analysis tools such as 
linear regression, random effects models and matrix decompositions. 
The `amen` package provides Bayesian model fitting algorithms
for AME models, and accommodates a variety of types of 
network relations, 
including continuous, binary and ordinal dyadic variables. 

The basic  AME model is of the form 
$y_{i,j} \sim  \beta^\top x_{i,j} + u_i^\top v_j + a_i +b_j + \epsilon_{i,j}$
where 

* $y_{i,j}$ is the observed dyadic variable being modeled
and $x_{i,j}$ is an observed vector of regressors;

* $a_i + b_j +  \epsilon_{i,j}$ is an  additive random effects term  that describes sender and receiver variance (such as outdegree and indegree 
heterogeneity) and dyadic correlation; 

* $u_i^\top v_j$ is a multiplicative random effects term 
that describes third-order 
dependence patterns (such as transitivity and clustering) and can be estimated 
and analyzed to uncover low-dimensional  structure in the network. 

#### Installation


```{r,eval=FALSE}
# Current version on GitHub
devtools::install_github("pdhoff/amen") 

# CRAN-approved version on CRAN
install.packages("amen")
```


#### Documentation

A tutorial and many data analysis examples are available 
via the [vignette](https://github.com/pdhoff/amen/blob/master/inst/doc/amen.pdf). Please cite this as 

Hoff, P.D. (2015) "Dyadic data analysis with *amen*". arXiv:1506.08237. 

#### Some examples

* [Modeling a binary outcome](articles/binary_demo.html)

* [Modeling an ordinal outcome](articles/ordinal_demo.html)

* [Modeling a continuous outcome](articles/continuous_demo.html)


